---
title: "Normal Model to Predict Housing Prices in Florida"
author: "Ethan Bensadoun"
date: "2024-08-15"
output: github_document
---

## Part 1

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(MASS)
library(ggrepel)
```

```{r}
Houses <- read_table("http://www.stat.ufl.edu/~aa/glm/data/Houses.dat")
# we can drop the X8 column (that is filled with NAs) by subsetting the Houses data set
Houses <- subset(Houses, select = -X8)
str(Houses)
print(Houses)
```

From the given data table on house data in Florida, has the following variables (for 100 observations) listed: - selling price (in thousands of dollars) - size of house (in square feet) - annual property tax bill (in dollars) - number of bedrooms - number of bathrooms - whether the house is new

According to section 3.4 in the textbook, the 100 observations are from one city alone, i.e, Gainseville. Thus, any inferences does not imply the relationships in general. They are treated as a random sample for illustrative purposes.

Here are the following summary statistics and plots:

## 3a)

```{r}
summary(Houses)
# Below we can see the the descriptive statistics for all 
# the different variables (i.e., columns) in the Houses dataset
```

```{r}
# For modelling y = selling price 
Houses_plot <- Houses
Houses_plot$new <- factor(Houses$new, levels = c(0, 1), labels = c("Old", "New"))


# Now plot, allowing ggplot2 to handle the coloring based on the factor levels
plot <- ggplot(Houses_plot, aes(x = size, y = price)) +
  geom_point(aes(color = new), stroke = 1, size = 1) +  
  # Use 'color' within aes for 'new' as factor
  ggtitle("Housing Prices vs House Sizes") + theme_minimal()
# Print the plot
print(plot)
# x-axis represents size 

ggplot(Houses, aes(x=taxes , y=price)) + 
  geom_point(colour="red",stroke=1) + 
  ggtitle("Housing Prices vs Taxes") + theme_minimal() 
# x-axis represents taxes 

ggplot(Houses, aes(x=beds , y=price)) + 
  geom_point(colour="pink",stroke=1) + 
  ggtitle("Housing Prices vs Number of Beds") + theme_minimal()
# x-axis represents beds 

ggplot(Houses, aes(x=baths , y=price)) + 
  geom_point(colour="purple",stroke=1) + 
  ggtitle("Housing Prices vs Number of Baths") + theme_minimal()
# x-axis represents baths 
```

## 3b)

```{r}
# We need to fit the model using the glm function 
House_fit <- glm(price ~  size + beds + baths + taxes, 
                 family=gaussian(link=identity), data=Houses_plot)
# The response variable follows a normal distribution, which is appropriate 
# for continuous outcome data.
# The identity link implies that changes in the predictors 
# have a direct and additive effect on the response variable's scale.
summary(House_fit)
```

The decision to choose this model is based on its fundamental nature as a linear model. The 'new' covariate is omitted from our analysis due to its binomial nature, which causes integration challenges when combined with other continuous or discrete non-binomial covariates in the modeling process. It would make more sense to run a logit regression model for binary covariates.

```{r}
# Next we can use the stepAIC() function with a forward direction 
# in order to select forward, starting with the null model, adding terms 
# in order to improve the fit. 
stepAIC(House_fit, direction = "forward")
# Notice that the case column is not included in the output, since it it just 
# an ascending list that is there to identify the housing unit.
```


We can interpret the result above from the House_fit and StepAIC summary output: 

- There seems to be a positive correlation between the price of a housing unit and size/taxes (as the coefficients are positive). 
- Thus, we can assume a correlation between increasing housing prices and and increase in size or taxes. - Specifically, a 1 unit rise in taxes correlates with a .040 unit increase in price, while a 1 unit rise in house size correlates with a .0746 unit increase in price.
- The positive correlation between taxes and price suggests that higher taxes are likely a consequence of higher house values, rather than being a direct value-adding factor.

- On the other hand, there seems to be a negative correlation between the price of a housing unit and beds/baths (as the coefficients are negative, -16.14218 and -1.34498). Thus, these covariates show a negative correlation with price, suggesting these factors might not always enhance house value as expected.
- Note, one could assume that an increase in bedrooms and bathrooms might lead to an increase in the housing unit's value, running stepAIC() identifies a negative correlation as previously mentioned. Thus, we can attribute this result to the wide value ranges for houses with the same number of beds and baths. Where some points might be weighted differently, negatively impacting the correlations for these covariates in the model.

Notice also that in the summary output table of House_fit, there are 3 stars on the right side of the p-values for size and taxes, which can be attributed by financial considerations in housing decisions (i.e., as the size if the house increases, the more the unit is taxed). This is a statistically significant result of the model.

## 3c)

```{r}
ggplot(data=Houses_plot,aes(y=residuals(House_fit),x=price)) + 
  geom_point(color = "red", aplha = 0.5, size=2) + 
  xlab("Housing Price") + ylab("Residuals") + 
  geom_hline(yintercept = 0) + theme_minimal()
```

- Notice the 3 outliers in the top right corner of the graph. Their position is considered to be highly influential indicating both high residuals and high prices. This discrepancy suggests that the model does not fully capture the pattern for these points, thus, they can disproportionally influence the model covariates as I previously mentioned in part(b). They can have more leverage, which inevitably affects the slope of the fitted model which skews model predictions about housing prices.
- In the summary output in part(b) we can see the null deviance is 1015000 (99 df) and residual deviance is 223400 (95 df). This reduction in deviance is good because it implies that adding a predictor provides a better fit. Thus, the covariates explain a significant portion of the variability in the housing price response variable. This improves the goodness of fit of the model.

## Part 2


```{r cars}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(MASS)
library(ggrepel)
library(ggfortify)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
Houses <- read_table("http://www.stat.ufl.edu/~aa/glm/data/Houses.dat")
# we can drop the X8 column (that is filled with NAs) by subsetting the Houses data set
Houses <- subset(Houses, select = -X8)
str(Houses)
print(Houses)
```

## 4a)

```{r}
Houses_fit2 <-glm(price ~ size + taxes + beds + baths, 
                  family=Gamma(log),data=Houses)
summary(Houses_fit2)
```

```{r}
Step_House <- stepAIC(Houses_fit2, direction = "forward")
summary(Step_House)
```

In the model above, we can interpret the coefficients of each covariate as follows:

-   Size: 3.269e-04, which indicates that for additional unit of size is associated with a 0.0327% increase in home value.
-   Taxes: 2.275e-04, which indicates that an increase in taxes paid by one unit results in a 0.0227% increase in home value.
-   Beds: -7.203e-02, which indicates that adding one bedroom to a house is linked to a 6.95% decrease in its value (because 1-.9305 = 0.0695).
-   Baths: 3.222e-02, which indicates that Each extra bathroom contributes to a 3.27% rise in home value.

The output also indicates that price, size and taxes are statistically significant (three stars). Where the 2 covariates have a p-value of 2e-16 and 0.000117. The other covariates (i.e., coefficients) won't have an intuitive correlation with home price since the p-values are not statistically siginificant, thus, we cannot say we are as fully confident with those results. There isn't evidence to say whether changing those variables will have a systematic effect on the housing price.

## 4.b)

```{r}
Houses_fit3 <-glm(price ~ size + taxes + beds + baths, 
                  family=Gamma(link = identity),data=Houses)
summary(Houses_fit3)
```

```{r}
Step_House2 <- stepAIC(Houses_fit3, direction = "forward")
summary(Step_House2)
```

For the adjusted model where link = identity, we can make the following interpretations of the coefficients:

-   Size: 0.065148, for an additional unit of size is associated with a 0.0651 units increase in home value.
-   Taxes: 0.038452, A one-unit increase in taxes paid boosts the home's value by approximately 0.0385 units.
-   Beds: -20.678022, adding a bedroom to the house reduces its value by about 20.68 units.
-   Baths: 9.585793, an additional bathroom raises the home's value by around 9.583 units.

Note there is statistical significance for the covariates size and taxes (sames as part a), with p-values 9.98e-07 and 1.53e-11. Note that beds is also statistically significant with a p-value of 0.00141. However, the covariate is not significant to the same degree as the other two.

Note that there is model has lower residual deviance of 6.6025 (95 df) whereas in (a), the residual deviance was 8.5167 (95 df). Thus, we can see an improvement in the performance of the statistical model in terms of how well it fits the data. They both outperform the intercept-only model which has a null deviance of 31.9401.

These findings are reminiscent of those in section (b) of Question 3, where the covariates'size' and 'taxes' were significant, and 'beds' showed some significance as well. Another point of comparison is that, despite not being statistically significant, the estimated effect of bedrooms on house price is the most substantial in this model and in section (b) of Question 3. That is, the addition of bedrooms contributes most significantly to changes in the house price, according to the estimates produced by these models using stepAIC().

## 4.c)

We can analyze the leverage of the models using the following functions:

```{r}
autoplot(Step_House2)
```

Residual vs Fitted Plot: the plot in the top left corner indicates a high variability in the residuals across the data spectrum, which implies the highest predicted prices. With a particular emphasis on points at the extreme right end.

Normal QQ Plot: It is definitely fitted as the points are on the line, however, there are some outliers that deviate from the line, which implies deviations from the fitted model.

Scale-Location Plot: We can see that the residuals seem to be more random as the line is more horizontal.

Residual vs Leverage Plot: It demonstrates that many data points exert a considerable influence on the model's outcomes, as they are bunched up to the right and near 0.

```{r}
autoplot(Step_House)
```

Residual vs Fitted Plot: the plot in the top left corner indicates a high variability in the residuals across the data spectrum, which implies the highest predicted prices. With a particular emphasis on points at the extreme right end.

Normal QQ Plot: It is definitely fitted as the points are on the line, however, there are some outliers that deviate from the line, which implies deviations from the fitted model. At the bottom on the left there seems to be more outliers that drop off from the fitted line.

Scale-Location Plot: We can see that the residuals seem to be more random as the line is more horizontal.

Residual vs Leverage Plot: The plot indicates that many data points significantly influence the model's predictions.

-   Lastly, we notice certain similarities from the model in q3. Specifically with the scale-location and residual plots. Specifically where the points bunch up to the left.
